The watershed transform in ITK - discussion and new developments

### 2005-11-16 Report/watershed_paper.tex
rjb - minor changes
darcs-hash:20051116061051-fafb9-8f8104079a241dcc835c618f022a50e685724608.gz

\begin{document}
\title{The watershed transform in ITK - discussion and new developments}
\author{Richard Beare{1} \and Gaetan Lehmann{2}}
\institute{Department of Medicine, Moansh University, Australia
\and Biologie du Developpement et de la Reproduction, INRA de Jouy-en-Josas (France)}
\maketitle

@@

{\em what about the topological watershed?}

@@

Early implementations used a special ordered queue for 8 bit data,
which may be where the misconception that a morphological watershed
can only be applied to 8 bit data types arose. The new ITK watershed
algorithms use standard c++ containers (maps and standard queues) to
implement the ordered queue\footnote{One potential problem with
applying this style of algorithm to non integer data is finding the
regional minima. Reconstruction approaches are often used to find
regional minima of integer type images, but subtracting 1, carrying
out a recursive dilation and subtracting the two. The same concept
does not extend to floating point data. However, if regional minima
are available the algorithms we are discussing can be applied to
floating point data.}.

@@

>>>>

\begin{document}
\title{The watershed transform in ITK - discussion and new developments}
\author{Richard Beare{$^1$} \and Gaetan Lehmann{$^2$}}
\institute{Department of Medicine, Moansh University, Australia
\and Biologie du Developpement et de la Reproduction, INRA de Jouy-en-Josas (France)}
\maketitle

@@

\subsection{The topological watershed}
An alternative approach, known as the topological watershed, has
recently been developed \cite{}. The topological watershed modifies
the topology of the surface so that all pixels that are not watershed
lines become regional minima and watershed lines are modified so that
they are more suitable for use in measuring saliency of boundaries
between regions. Algorithms implementing the topological watershed
also exploit priority queues, but don't appear to be able to avoid the
need for explicit lower completion or reconstruction by erosion.

I don't understand the implications of the topological watershed yet.

@@

Early implementations used a special ordered queue for 8 bit data,
which may be where the misconception that a morphological watershed
can only be applied to 8 bit data types arose -- in fact some of the
early work along these lines was covered by a patent\footnote{I'm told
that the associated paper is \cite{Meyer91a} and that there may be an
English version in \cite{Dougherty93a}}. The new ITK watershed
algorithms use standard c++ containers (maps and standard queues) to
implement the ordered queue\footnote{One potential problem with
applying this style of algorithm to non integer data is finding the
regional minima. Reconstruction approaches are often used to find
regional minima of integer type images, but subtracting 1, carrying
out a recursive dilation and subtracting the two. The same concept
does not extend to floating point data. However, if regional minima
are available the algorithms we are discussing can be applied to
floating point data.}.

@@

\section{Finding regional minima}
The marker based watershed filter can be used as a convention
watershed algorithm by finding the regional minima and using them as
markers. Regional minima (and maxima) are often useful, and there are
a number of different ways of finding them:
\begin{itemize}
\item Using reconstruction approaches mentioned earlier - 
the problem with these is that a value of 1 is subtracted from the
original to produce the image from which the reconstruction is carried
out, which makes the approach less useful for floating point
types. The advantage of this approach is that it can be easily
extended to find {\em image dynamics}, which are often more stable and
less subject to noise than regional minima.
\item The way ITK does it, which is related to the connected 
component labeller. Finds all flat regions and then figures out which
are plateaus, minima and maxima by looking at the boundary values.
\item Flooding transform (not sure where this came from) which 
sets anything not a regional minima to a specified max value. The
algorithm assumes that the border is set to the maximum value. Pixels
are visited in raster order. If a pixel $A$ has a neighbour $B$ and
$P(A) > p(B)$ (i.e grey level of A is greater than B) then all pixels
connected to $A$ with value $P(A)$ are set to the maximum value. This
is a recursive algorithm that finds pixels that cannot be a regional
minimum (because they have a lower neighbour) and then eliminates any
connected region that also cannot be a regional minima.

The result would then need to be thresholded and labelled to produce
markers. This approach is not limited by the pixel type, but does
require a more random visit order when flooding, which could restrict
performance on larger images.
\end{itemize}
